Towards a Verified Tableau Prover for Multi-Level-Syllogistics

### 2022-08-23 paper/paper.tex
Continue

@@

\begin{abstract}
  TODO
\end{abstract}

@@

\section{Introduction}
TODO

@@

\subsection{Contributions}
TODO

\subsection{Notation}
Isabelle/HOL~\cite{isabelle} conforms to everyday mathematical notation for the most part.
For the benefit of the reader that is unfamiliar with Isabelle/HOL, we establish notation and in particular some essential datatypes together with their primitive operations that are specific to Isabelle/HOL.
We write \lstinline!t :: 'a! to specify that the term \lstinline!t! has the type \lstinline!'a! and \lstinline!'a $\Rightarrow$ 'b! for the type of a total function from \lstinline!'a! to \lstinline!'b!.
The types for booleans, natural numbers, and in integers are \lstinline!bool!, \lstinline!nat!, and \lstinline!int!, respectively.
Sets with elements of type \lstinline!'a! have the type \lstinline!'a set!.
Analogously, we use \lstinline!'a list! to describe lists, which are constructed as the empty list \lstinline![]! or with the infix constructor \lstinline!#!, and are appended with the infix operator \lstinline!@!.
The function \lstinline!set! converts a list into a set.
For optional values, Isabelle/HOL offers the type \lstinline!option! where a term \lstinline!opt :: 'a option! is either \lstinline!None! or \lstinline!Some a! with \lstinline!a :: 'a!.
Finally, we remark that \textbf{iff} is equivalent to \lstinline!$=$! on type \lstinline!bool! and \lstinline!$\equiv$! is definitional equality of the meta-logic of Isabelle/HOL, which is called Isabelle/Pure.
TODO: Add arrows and bracket notation for Pure formulae
\section{A Semantics for Sets\label{sec:semantics}}
\begin{lstlisting}
datatype 'a fm =
  Atom 'a | Neg ('a fm)
| And ('a fm) ('a fm) | Or ('a fm) ('a fm)
\end{lstlisting}

% TODO: Think about introducing notation for single
\begin{lstlisting}
datatype 'a pset_term =
  $\emptyset$ | Var 'a | Single ('a pset_term)
| 'a pset_term $\sqcup_\text{s}$ 'a pset_term
| 'a pset_term $\sqcap_\text{s}$ 'a pset_term
| 'a pset_term $-_\text{s}$ 'a pset_term
\end{lstlisting}

\begin{lstlisting}
datatype 'a pset_atom =
  'a pset_term $\in_\text{s}$ 'a pset_term
| 'a pset_term $\approx_\text{s}$ 'a pset_term
\end{lstlisting}

\begin{lstlisting}
type_synonym 'a pset_fm = 'a pset_atom fm
type_synonym 'a branch = 'a pset_fm list
\end{lstlisting}

\begin{lstlisting}
definition vars_fm :: 'a pset_fm $\Rightarrow$ 'a set where
  vars_fm $\phi$ $\equiv$ $\bigcup$(set_pset_atom ` atoms $\phi$)
\end{lstlisting}

@@

\begin{lstlisting}
fun I$_\text{st}$ :: ('a $\Rightarrow$ V) $\Rightarrow$ 'a pset_term $\Rightarrow$ V where
  I$_\text{st}$ v $\emptyset$
| I$_\text{st}$ v (Var x) = v x
| I$_\text{st}$ v (Single s) = vset {I$_\text{st}$ v s}
| I$_\text{st}$ v (s $\sqcup_\text{s}$ t) = I$_\text{st}$ v s $\sqcup$ I$_\text{st}$ v t 
| I$_\text{st}$ v (s $\sqcap_\text{s}$ t) = I$_\text{st}$ v s $\sqcap$ I$_\text{st}$ v t 
| I$_\text{st}$ v (s $-_\text{s}$ t) = I$_\text{st}$ v s $-$ I$_\text{st}$ v t 
\end{lstlisting}

\begin{lstlisting}
fun I$_\text{sa}$ :: ('a $\Rightarrow$ V) $\Rightarrow$ 'a pset_atom $\Rightarrow$ V where
  I$_\text{sa}$ v (s $\in_\text{s}$ t) $\longleftrightarrow$ I$_\text{st}$ v s $\in$ elts (I$_\text{st}$ v t)
| I$_\text{sa}$ v (s $\approx_\text{s}$ t) $\longleftrightarrow$ I$_\text{st}$ v s = I$_\text{st}$ v t
\end{lstlisting}

\begin{lstlisting}
fun subterms :: 'a pset_term $\Rightarrow$ 'a pset_term set
where
  subterms $\emptyset$ = {$\emptyset$}
| subterms (Var x) = {Var x}
| subterms (s $\sqcup_\text{s}$ t) = {s $\sqcup_\text{s}$ t}
    $\cup$ subterms s $\cup$ subterms t
| subterms (s $\sqcap_\text{s}$ t) = {s $\sqcap_\text{s}$ t}
    $\cup$ subterms s $\cup$ subterms t
| subterms (s $-_\text{s}$ t) = {s $-\text{s}$ t}
    $\cup$ subterms s $\cup$ subterms t
| subterms (Single t) = {Single t} $\cup$ subterms t
\end{lstlisting}

\begin{lstlisting}
fun subterms_atom ::
      'a pset_atom $\Rightarrow$ 'a pset_term set where
  subterms_atom (s $\in_\text{s}$ t) = subterms s $\cup$ subterms t
| subterms_atom (s $\approx_\text{s}$ t) = subterms s $\cup$ subterms t
\end{lstlisting}

\begin{lstlisting}
definition subterms_fm ::
    'a pset_fm $\Rightarrow$ 'a pset_term set where
  subterms_fm $\phi$ $\equiv$ $\bigcup$(subterms_atom ` atoms $\phi$)
\end{lstlisting}

\section{A Tableau Calculus for MLSS}
Interesting: they forgot one rule.

@@

@@

\subsection{Termination}
Maybe interesting: technical lemma that allows to obtain the point at which a parameter was introduced.
\subsection{Soundness}
Interesting: The model changes because auf new parameters.
Maybe move this to the chapter on the calculus?
\subsection{Lifting the Completeness of the Calculus}
TODO

@@

\section{Conclusion}
TODO

\begin{acks}
  TODO
\end{acks}

>>>>

\usepackage{todonotes}

@@

\begin{abstract}
  \todo[inline]{Write the abstract}
\end{abstract}

@@

\listoftodos{}

\section{Introduction}
\todo[inline]{Write the introduction}

@@

\subsection{Contributions}
\todo[inline]{Write the contributions}

\subsection{Notation}
Isabelle/HOL~\cite{isabelle} conforms to everyday mathematical notation for the most part.
For the benefit of the reader that is unfamiliar with Isabelle/HOL, we establish notation and in particular some essential datatypes together with their primitive operations that are specific to Isabelle/HOL.
We write \lstinline!t :: 'a! to specify that the term \lstinline!t! has the type \lstinline!'a! and \lstinline!'a $\Rightarrow$ 'b! for the type of a total function from \lstinline!'a! to \lstinline!'b!.
The types for booleans, natural numbers, and in integers are \lstinline!bool!, \lstinline!nat!, and \lstinline!int!, respectively.
Sets with elements of type \lstinline!'a! have the type \lstinline!'a set!.
Analogously, we use \lstinline!'a list! to describe lists, which are constructed as the empty list \lstinline![]! or with the infix constructor \lstinline!#!, and are appended with the infix operator \lstinline!@!.
The function \lstinline!set! converts a list into a set.
For optional values, Isabelle/HOL offers the type \lstinline!option! where a term \lstinline!opt :: 'a option! is either \lstinline!None! or \lstinline!Some a! with \lstinline!a :: 'a!.
Finally, we remark that \textbf{iff} is equivalent to \lstinline!$=$! on type \lstinline!bool! and \lstinline!$\equiv$! is definitional equality of the meta-logic of Isabelle/HOL, which is called Isabelle/Pure.
\todo[inline]{Add arrows and bracket notation for Pure formulae}
\section{Syntax and Semantics of MLSS\label{sec:semantics}}
\subsection{Syntax}
At the heart of \textbf{MLSS}, we have the type of set terms which includes the empty set and variables as well as the operations union, intersection, set difference, and \lstinline!Single!, which represents a singleton set containing the argument.
\todo[inline]{Syntax for \lstinline!Single!?}
\begin{lstlisting}
datatype 'a pset_term =
  $\emptyset$ | Var 'a | Single ('a pset_term)
| 'a pset_term $\sqcup_\text{s}$ 'a pset_term
| 'a pset_term $\sqcap_\text{s}$ 'a pset_term
| 'a pset_term $-_\text{s}$ 'a pset_term
\end{lstlisting}
We can combine two set terms to form a set atom by using the membership or the equality operator.
\begin{lstlisting}
datatype 'a pset_atom =
  'a pset_term $\in_\text{s}$ 'a pset_term
| 'a pset_term $\approx_\text{s}$ 'a pset_term
\end{lstlisting}
With the above operators we can also represent the subset operator \lstinline!$\sqsubseteq_\text{s}$! and enumerate finite sets as \lstinline!$s \subseteq_\text{s} t$! is equivalent to \lstinline!s $\sqcup_\text{s}$ t $\approx_\text{s}$ t! and a finite set of elements \lstinline!{t$_1$,t$_2$,$\ldots$,t$_k$}! can be expressed by \lstinline!Single t$_1$ $\sqcup_\text{s}$ $\ldots$ $\sqcup_\text{s}$ Single t$_k$!.

The unquantified fragment \textbf{MLSS} of set theory is then formed by combining set atoms with propositional logical connectives.
\begin{lstlisting}
datatype (atoms: 'a) fm =
  Atom 'a | Neg ('a fm)
| And ('a fm) ('a fm) | Or ('a fm) ('a fm)

type_synonym 'a pset_fm = 'a pset_atom fm
\end{lstlisting}
Note that Isabelle's datatype package automatically defines a function \lstinline!atoms :: 'a fm $\Rightarrow$ 'a set! that allows us to obtain all set atoms in the formula.
Similarly, it defines a function \lstinline!set_pset_atom :: 'a pset_atom $\Rightarrow$ 'a set! with which we can extract all the variable names in a set atom.
We combine these functions in order to define \lstinline!vars_fm! that extracts all the variable names occuring in a set formula.
\begin{lstlisting}
definition vars_fm :: 'a pset_fm $\Rightarrow$ 'a set where
  vars_fm $\phi$ $\equiv$ $\bigcup$(set_pset_atom ` atoms $\phi$)
\end{lstlisting}
In addition to the variable names that appear in a formula, we will later also need to obtain those set terms that occur as subterms in a formula. 
We first define a function that collects the subterms of a set term.
\begin{lstlisting}
fun subterms ::
    'a pset_term $\Rightarrow$ 'a pset_term set where
  subterms $\emptyset$ = {$\emptyset$}
| subterms (Var x) = {Var x}
| subterms (s $\sqcup_\text{s}$ t) =
    {s $\sqcup_\text{s}$ t} $\cup$ subterms s $\cup$ subterms t
| subterms (s $\sqcap_\text{s}$ t) =
    {s $\sqcap_\text{s}$ t} $\cup$ subterms s $\cup$ subterms t
| subterms (s $-_\text{s}$ t) =
    {s $-_\text{s}$ t} $\cup$ subterms s $\cup$ subterms t
| subterms (Single t) =
    {Single t} $\cup$ subterms t
\end{lstlisting}
The subterms of an atom are then the subterms of those terms that make up the atom.
\begin{lstlisting}
fun subterms_atom ::
      'a pset_atom $\Rightarrow$ 'a pset_term set where
  subterms_atom (s $\in_\text{s}$ t) =
    subterms s $\cup$ subterms t
| subterms_atom (s $\approx_\text{s}$ t) =
    subterms s $\cup$ subterms t
\end{lstlisting}
Ultimately, we lift this to formulae by using \lstinline!atoms! again.
\begin{lstlisting}
definition subterms_fm ::
    'a pset_fm $\Rightarrow$ 'a pset_term set where
  subterms_fm $\phi$ $\equiv$ $\bigcup$(subterms_atom ` atoms $\phi$)
\end{lstlisting}
The last function that we define computes the subformulae of a formula.
\begin{lstlisting}
fun subfms :: 'a fm $\Rightarrow$ 'a fm set where
  subfms (Atom a) = {Atom a}
| subfms (And p q) =
    {And p q} $\cup$ subfms p $\cup$ subfms q
| subfms (Or p q) =
    {Or p q} $\cup$ subfms p $\cup$ subfms q
| subfms (Neg q) = {Neg q} $\cup$ subfms q
\end{lstlisting}

\subsection{Semantics}
Formulae in \textbf{MLSS} contain arbitrarily nested set terms.
We account for this arbitrary nesting by basing the semantics on the von Neumann hierarchy $\mathcal{V}$ of sets that are defined inductively as
  \[
    \begin{array}{rclr}
      \mathcal{V}_0 & = & \emptyset, \\
      \mathcal{V}_{\alpha + 1} & = & \mathcal{P}(\mathcal{V}_\alpha) & \text{for each ordinal $\alpha$}, \\
      \mathcal{V}_{\lambda} & = & \bigcup_{\mu < \lambda} \mathcal{V}_\mu & \text{for each limit ordinal $\lambda$}, \\
      \mathcal{V} & = & \bigcup_\alpha \mathcal{V}_\alpha, \\
    \end{array}
  \]
where $\mathcal{P}(S)$ is the powerset of $S$.
The sets in $\mathcal{V}$ are well-founded meaning that there can be no membership cycle in $\mathcal{V}$.

In Isabelle/HOL, which is simply typed, this definition is not accepted, though.
A way to work around this limitiation is provided by Lawrence in an AFP entry~\cite{zfc_in_hol_afp}, which adds $\mathcal{V}$ to Isabelle/HOL by way of axiomatisation.
An article by Lawrence~\cite{zfc_in_hol_afp} provides some further context to this formalisation.

\begin{lstlisting}
fun I$_\text{st}$ :: ('a $\Rightarrow$ V) $\Rightarrow$ 'a pset_term $\Rightarrow$ V where
  I$_\text{st}$ M $\emptyset$
| I$_\text{st}$ M (Var x) = M x
| I$_\text{st}$ M (Single s) = vset {I$_\text{st}$ M s}
| I$_\text{st}$ M (s $\sqcup_\text{s}$ t) = I$_\text{st}$ M s $\sqcup$ I$_\text{st}$ M t 
| I$_\text{st}$ M (s $\sqcap_\text{s}$ t) = I$_\text{st}$ M s $\sqcap$ I$_\text{st}$ M t 
| I$_\text{st}$ M (s $-_\text{s}$ t) = I$_\text{st}$ M s $-$ I$_\text{st}$ M t 
\end{lstlisting}

\begin{lstlisting}
fun I$_\text{sa}$ :: ('a $\Rightarrow$ V) $\Rightarrow$ 'a pset_atom $\Rightarrow$ V where
  I$_\text{sa}$ M (s $\in_\text{s}$ t) $\longleftrightarrow$ I$_\text{st}$ M s $\in$ elts (I$_\text{st}$ M t)
| I$_\text{sa}$ M (s $\approx_\text{s}$ t) $\longleftrightarrow$ I$_\text{st}$ M s = I$_\text{st}$ M t
\end{lstlisting}

\section{A Tableau Calculus for MLSS}
\begin{lstlisting}
type_synonym 'a branch = 'a pset_fm list
\end{lstlisting}

@@

Interesting: they forgot one rule.

@@

\begin{lstlisting}
definition wf_branch b $\equiv$ $\exists$$\phi$. extendss b [$\phi$]
\end{lstlisting}

\begin{lstlisting}
definition params :: 'a branch $\Rightarrow$ 'a set where
  params b $\equiv$ vars_branch b - vars_fm (last b)
\end{lstlisting}

\begin{lstlisting}
definition params' :: 'a branch $\Rightarrow$ 'a set where
  params' b $\equiv$
    {c $\in$ params b. $\forall$t $\in$ subterms_fm (last b).
      AT (Var c $\approx_\text{s}$ t) $\notin$ set b $\land$
      AT (t $\approx_\text{s}$ Var c) $\notin$ set b} 
\end{lstlisting}

\begin{lstlisting}
definition subterms_branch' ::
    'a branch $\Rightarrow$ 'a pset_term set where
  subterms_branch' b $\equiv$
    subterms_fm (last b) $\cup$ 
    Var ` (params b - params' b)
\end{lstlisting}

\begin{lstlisting}
definition bgraph $\equiv$ $\ldots$ 
\end{lstlisting}

\begin{lstlisting}
definition no_new_subterms b $\equiv$
  $\forall$t $\in$ subterms_branch b. t $\notin$ Var ` params b
    $\longrightarrow$ t $\in$ subterms_fm (last b))
\end{lstlisting}

@@

\subsection{Termination}
Maybe interesting: technical lemma that allows to obtain the point at which a parameter was introduced.
\subsection{Soundness}
Interesting: The model changes because auf new parameters.
Maybe move this to the chapter on the calculus?
\subsection{Lifting the Completeness of the Calculus}
\todo[inline]{Do you even lift?}

@@

\section{Conclusion}
\todo[inline]{Conclude}

\begin{acks}
  \todo[inline]{Acknowledge}
\end{acks}
